---
title: "Milestone #4"
author: "Rachael Baartmans, Lara Petalio, Christine Truong"
date: "11-21-22"
output:
  pdf_document: default
  html_document: default
---
```{r previous import and data cleaning steps, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(knitr)
library(kableExtra)

#Insert previous code with updated datasets 
race_data <-read_csv("ca_csc_outcome_race_data.csv",
            col_select = c(ID, NERVOUS, WORRYING, PROBINTR,
                           PROBDOWN, ASTHMA, HEARTDIS,
                           DIABETES, OTHMENILL, race01, race02, race03,
                           race04, race05, race06, race07, race08,
                           race09, race10, race11, race12, race13,
                           race14, race15),
            na = c("", "NA", "NA/Not Applicable", "N/A", "n/a",
                   "(DO NOT READ) NA/Not Applicable",
                   "(DO NOT READ) Refused",
                   "(DO NOT READ) Don't know"))

smoker_data <- read_csv("ca_csc_smoker_data.csv",
            col_select = c(psraid, smokstat, WHEREBUY, BUYCALIF,
                           HOWMANY, SMOK6NUM, SMOK6UNI),
            na = c("", "NA", "NA/Not Applicable", "N/A", "n/a",
                   "(DO NOT READ) NA/Not Applicable",
                   "(DO NOT READ) Refused",
                   "(DO NOT READ) Don't know"))

#Changed casing for variables from capitals to lowercase in both dataframes
#of race_data and smoker_data
names(race_data) <- tolower(names(race_data))
names(smoker_data) <- tolower(names(smoker_data))

#Re-coded "100 or more cigarettes" to "100" for future pack-year calculations
#once the variable `howmany` is converted from character to numeric data type
smoker_data$howmany <- recode(smoker_data$howmany,
                              "100 or more cigarettes" = "100")

#Changed the data type of `howmany` from character to numeric in order to
#perform calculations for pack-years later
smoker_data$howmany <- as.numeric(smoker_data$howmany)

#Re-coded "In military commissaries, or" to "In military commissaries", as well
#as "Somewhere else (SPECIFY)?" to "Somewhere else" to make
#response option more understandable when displayed for the variable `wherebuy`.
smoker_data$wherebuy <- recode(smoker_data$wherebuy,
                  "In military commissaries, or" = "In military commissaries",
                  "Somewhere else (SPECIFY)?" = "Somewhere else")

#Created new variable `race` to combine variables race01:race15 into one column. 
race_data_2 <- race_data %>%
  mutate(race = case_when(race01 == "Yes" ~ "White",
        race02 == "Yes" ~ "Black",
        race03 == "Yes" ~ "Japanese",
        race04 == "Yes" ~ "Chinese",
        race05 == "Yes" ~ "Filipino",
        race06 == "Yes" ~ "Korean",
        race07 == "Yes" ~ "Other Asian or Pacific Islander",
        race08 == "Yes" ~ "American Indian or Alaskan Native",
        race09 == "Yes" ~ "Mexican",
        race10 == "Yes" ~ "Hispanic/Latino",
        race11 == "Yes" ~ "Other",
        race12 == "Yes" ~ "Vietnamese",
        race13 == "Yes" ~ "Asian Indian",
        race14 == "Yes" ~ "Refused",
        race15 == "Yes" ~ "Don't know")) %>%
  select(-(race01:race15))

#Created new variable "packs_per_day" for future calculations for pack-years
smoker_data_2 <- smoker_data %>% mutate(packs_per_day = howmany/20)

```
  
```{r pack-years calculation, echo=F}
#We calculated pack-years, which is given by the formula of
#pack-years = # of packs of cigarettes smoked per day * years a person has smoked.
#This calculation led to the creation of a new variable called `pack_years`.
#`pack_years` was created conditionally based on the three different time units
#as determined by the existing variable `smok6uni`, which are: "Days", "Months",
#and "Years". To change the unit of "Days" to years, we set up a conditional
#statement in the code to divide `smok6num` by 365 before multiplying the result
#by `packs_per_day` to get pack-years. Similarly, to change the unit of "Months"
#to years, we set up a conditional statement in the code to divide `smok6num`
#by 12 before multiplying the result by `packs_per_day` to get pack-years. For
#`smok6uni` observations that have the value "Years", we just multiplied
#`smok6num` by `packs_per_day` to get pack-years directly. We assigned
#this overall change in the data frame smoker_data_2 to a new data frame
#called smoker_data_3, which includes the new variable `pack_years`.

smoker_data_3 <- smoker_data_2 %>%
  mutate(pack_years =
           case_when(smok6uni == "Days" ~ packs_per_day*(smok6num/365),
                     smok6uni == "Months" ~ packs_per_day*(smok6num/12),
                     smok6uni == "Years" ~ packs_per_day*(smok6num)))

#We then rounded `pack_years` to the nearest whole number for all observations.

smoker_data_3$pack_years <- round(smoker_data_3$pack_years, 0)

```

```{r joining race_data_2 and smoker_data_3 together, echo=F}

#In order to join our two cleaned data sets together, we first had to remove the
#strings of 'DIS' and 'STAT' from the `id` column of race_data_2 by using gsub().
#We overwrote these changes in the race_data_2 data frame and viewed these new
#changes to make sure the `id` variable only contains numbers and no
#characters.

race_data_2$id <- gsub('[DISSTAT]', '', race_data_2$id)

#Next, looking at the smoker_data_3 data frame, we see that the `psraid`
#variable contains each study participant's unique ID number, but the variable
#is a numeric data type. On the other hand, `id` from the race_data_2
#data frame is a character data type. We needed to convert `psraid` then from
#character to numeric data type because 1) `psraid` is an identifier rather than
#a numeric value to mathematically manipulate even if it does contain numbers
#and 2) in order to perform a join, the two variables must be the same data
#type.

smoker_data_3$psraid <- as.character(smoker_data_3$psraid)

#Afterward, we performed an inner join between race_data_2 and
#smoker_data_3 by each study participant's unique ID number, which is
#represented by `id` in race_data_2 and `psraid` in smoker_data_3. We
#chose to do an inner join because we wanted to select participants that
#exist in each of our two data sets for our final data frame.
#We assigned this join to a new data frame called joined_smoking_df.

joined_smoking_df <- inner_join(x = race_data_2, y = smoker_data_3,
                                by=c("id" = "psraid"))

```
## _NOTE: We calculated pack-years and joined our two data sets of race_data_2 and smoker_data_3 together before creating these visualizations below. The pack-years calculation was created as the new variable `pack_years` in our joined data set, joined_smoking_df. If necessary, please see the code chunks labeled "r joining race_data_2 and smoker_data_3 together" and "r pack-years calculation" in Milestone_4.Rmd to view these processes not mentioned in this pdf._

\newpage

## Visualizations

_**Table: Average Number of Pack-years by Disease Outcome Among Smokers**_
```{r table of cigarette pack-years by disease outcome among smokers}

#Table for average number of pack-years per disease for smokers who have a disease
t_avg_pack_years_disease <- joined_smoking_df %>%
  mutate(disease = case_when(asthma == "Yes" ~ "Asthma",
                            heartdis == "Yes" ~ "Heart Disease",
                            diabetes == "Yes" ~ "Diabetes",
                            othmenill == "Yes" ~ "Mental Illness")) %>%
  select(disease, pack_years) %>%
  filter(!is.na(pack_years), !is.na(disease)) %>%
  group_by(disease) %>%
  summarize(avg_pack_years = round(sum(pack_years)/n(), 0))


#Kable table for average number of pack-years per disease outcome for smokers who
#have a disease (produced below)
kable(t_avg_pack_years_disease, 
booktabs=T, 
col.names=c("Disease", "Average Number of Pack-years"),  
align='lcccc', 
caption= 'Average Number of Pack-years by Disease Outcome Among Smokers') %>%
kable_styling(full_width = T) %>%
kable_styling(latex_options = "hold_position") %>%
footnote(general = 
           "Data Source: 2011 California Smokers Cohort, CA Dept. of Health")

```
_**Interpretation of Average Number of Pack-years by Disease Outcome Among Smokers Table:**_
\newline
This table demonstrates the average number of pack-years per disease type for smokers who reported having asthma, diabetes, heart disease, and/or mental illness in the 2011 California Smokers Cohort study.

Among smokers who have reported having asthma, heart disease, diabetes, and/or mental illness, those with heart disease have the highest number of average pack-years (28), while those with mental illness have the lowest number of average pack-years (17).

\newpage

_**Bar Graph: Average Number of Pack-years by Race and Mental Illness Status Among Smokers**_
```{r bar graph of cigarette pack-years by race and mental illness among smokers, fig.width=8, fig.height=5, message=FALSE, warning=FALSE}

#We first created a subset of the data frame joined_smoking_df for our disease of
#interest, mental illness, called avg_pack_years_race_othmenill. This subset
#includes only the variable of `race` and average values of the variable
#`pack_years` pertaining to mental illness status. The purpose of creating this
#subset is to simplify the process of creating a graph in the next step by
#showing only the relevant information we need.

avg_pack_years_race_othmenill <- joined_smoking_df %>%
  filter(!is.na(pack_years)) %>%
  group_by(race, othmenill) %>%
  summarize(avg_pack_years = sum(pack_years)/n())


#We then created a bar graph representing avg_pack_years_race_othmenill,
#excluding NA values in the variable `othmenill` since we have determined that
#the NA values do not present valuable information for our analyses; the NA
#values had already been dropped for `avg_pack_years` in the process of creating
#the subset of avg_pack_years_race_othmenill in the previous step.

avg_pack_years_race_othmenill %>%
  drop_na(othmenill) %>%
  ggplot(aes(x = race, y = avg_pack_years)) +
  geom_bar(aes(fill = race), stat = "identity", position = "dodge") +
  coord_flip() +
  guides(fill = "none") +
  theme(plot.title.position = "plot",
        plot.title = element_text(hjust = 0.5)) +
  labs(x = "Race",
       y = "Average Number of Pack-years",
  title = "Average Number of Pack-years by Race & Mental Illness Status Among Smokers",
  caption = "Data Source: 2011 California Smokers Cohort, CA Dept. of Health") +
  facet_wrap(~ othmenill, labeller = labeller(othmenill =
                                      c("No" = "Does Not Have Mental Illness",
                                        "Yes" = "Has Mental Illness"))) +
  theme(plot.title = element_text(hjust = 0.5),
        plot.caption = element_text(hjust = 0.5))
       
```
\newline 
\newline 
\newline _**Interpretation of Average Pack-years by Race and Mental Illness Status Among Smokers Bar Graph:**_
\newline
\newline
This graph exhibits the number of average pack-years for each race category and by mental illness status of smokers in the 2011 California Smokers Cohort study.

Among smokers who have reported having no mental illness, those who identified as "Hispanic/Latino" by race appear to have the greatest number of average pack-years, followed by "White and "American Indian or Alaskan Native", out of all race categories in the 2011 California Smokers Cohort.

Among smokers who have reported having mental illness, those who identified as "Other" by race appear to have the greatest number of average pack-years compared to other races in the 2011 California Smokers Cohort, with "White" and "American Indian or Alaskan Native" following closely behind.

\newpage

_**Bar Graph: Cigarette-buying Location Frequencies by Mental Illness Status Among Smokers**_
```{r bar graph of frequency per cigarette purchase location by mental illness among smokers, fig.width=8, fig.height=5}

#As similarly performed for the graph above, we also excluded NA values for the
#variables of `wherebuy` and `othmenill` prior to creating this graph because
#we did not believe NA values would be telling us any valuable information.
#We chose to create a stacked bar graph instead of a dodged bar graph in order
#to facilitate total frequency comparisons between different cigarette purchase
#locations regardless of mental illness status.

joined_smoking_df %>%
  filter(!is.na(wherebuy), !is.na(othmenill)) %>%
  ggplot(aes(x = wherebuy)) + 
  geom_bar(aes(fill = othmenill), position = "stack") +
  coord_flip() +
  theme(plot.title.position = "plot",
        plot.title = element_text(hjust = 0.5)) +
  scale_fill_discrete(name = "Has Mental Illness") + 
  labs(x = "Location of Cigarette Purchase",
       y = "Frequency",
  title = "Cigarette-buying Location Frequencies by Mental Illness Status Among Smokers",
  caption = "Data Source: 2011 California Smokers Cohort, CA Dept. of Health")

```
\newpage

_**Interpretation of Cigarette-buying Location Frequencies by Mental Illness Status Among Smokers Bar Graph:**_
\newline
\newline
This bar graph explores the relationship between frequencies per cigarette purchase location and mental illness status among smokers in the 2011 California Smokers Cohort study.

Mental illness was not reported by the majority of the smokers for each cigarette purchase location. However, mental illness was reported in the greatest number by those who purchased cigarettes at convenience stores or gas stations, followed by those who purchased cigarettes at tobacco discount stores; these are the two locations that also have the highest frequencies among smokers for making cigarette purchases at.
