---
title: "Milestone #3"
author: "Rachael Baartmans, Lara Petalio, Christine Truong"
date: "Sys.Date()"
output:
  pdf_document: default
  html_document: default
---
```{r importing data}
library(readr)

race_data <-read_csv("ca_csc_outcome_race_data.csv",
            col_select = c(NERVOUS, WORRYING, PROBINTR,
                           PROBDOWN, ASTHMA, HEARTDIS,
                           DIABETES, OTHMENILL, race01, race02, race03,
                           race04, race05, race06, race07, race08,
                           race09, race10, race11, race12, race13,
                           race14, race15),
            na = c("", "NA", "NA/Not Applicable", "N/A", "n/a",
                   "(DO NOT READ) NA/Not Applicable",
                   "(DO NOT READ) Refused",
                   "(DO NOT READ) Don't know"))

smoker_data <- read_csv("ca_csc_smoker_data.csv",
            col_select = c(smokstat, WHEREBUY, BUYCALIF,
                           HOWMANY, SMOK6NUM, SMOK6UNI),
            na = c("", "NA", "NA/Not Applicable", "N/A", "n/a",
                   "(DO NOT READ) NA/Not Applicable",
                   "(DO NOT READ) Refused",
                   "(DO NOT READ) Don't know"))
```
\newpage

## Subset rows and columns as needed 
```{r subset rows/columns}

```

\newpage
## Create New Variables needed for analysis
**Minimum of 2 created from existing columns **
\newline**Examples: calculating the rate or combining character strings**

```{r new variables}

```

\newpage
## Clean variables for analysis 
**Minimum of 2**
\newline **Examples: Recode invalid values/handle missing fields/recode categories **

```{r clean variables}

```

\newpage
## Data dictionary based on clean dataset 
**must include: variable name, data type, and description**
```{r data dictionary}

```

\newpage
## Tables with descriptive statistics for 4 data elements 
```{r tables}

```

