---
title: "Milestone #4"
author: "Rachael Baartmans, Lara Petalio, Christine Truong"
date: "11-14-22"
output:
  pdf_document: default
  html_document: default
---
```{r previous import and data cleaning steps, echo=FALSE}
library(tidyverse)
library(knitr)
library(kableExtra)

#Insert previous code with updated datasets 
race_data <-read_csv("ca_csc_outcome_race_data.csv",
            col_select = c(ID, NERVOUS, WORRYING, PROBINTR,
                           PROBDOWN, ASTHMA, HEARTDIS,
                           DIABETES, OTHMENILL, race01, race02, race03,
                           race04, race05, race06, race07, race08,
                           race09, race10, race11, race12, race13,
                           race14, race15),
            na = c("", "NA", "NA/Not Applicable", "N/A", "n/a",
                   "(DO NOT READ) NA/Not Applicable",
                   "(DO NOT READ) Refused",
                   "(DO NOT READ) Don't know"))

smoker_data <- read_csv("ca_csc_smoker_data.csv",
            col_select = c(psraid, smokstat, WHEREBUY, BUYCALIF,
                           HOWMANY, SMOK6NUM, SMOK6UNI),
            na = c("", "NA", "NA/Not Applicable", "N/A", "n/a",
                   "(DO NOT READ) NA/Not Applicable",
                   "(DO NOT READ) Refused",
                   "(DO NOT READ) Don't know"))

#Changed casing for variables from capitals to lowercase in both dataframes
#of race_data and smoker_data
names(race_data) <- tolower(names(race_data))
names(smoker_data) <- tolower(names(smoker_data))

#Re-coded "100 or more cigarettes" to "100" for future pack-year calculations
#once the variable `howmany` is converted from character to numeric data type
smoker_data$howmany <- recode(smoker_data$howmany,
                              "100 or more cigarettes" = "100")

#Changed the data type of `howmany` from character to numeric in order to
#perform calculations for pack-years later
smoker_data$howmany <- as.numeric(smoker_data$howmany)

#Re-coded "In military commissaries, or" to "In military commissaries", as well
#as "Somewhere else (SPECIFY)?" to "Somewhere else" to make
#response option more understandable when displayed for the variable `wherebuy`.
smoker_data$wherebuy <- recode(smoker_data$wherebuy,
                  "In military commissaries, or" = "In military commissaries",
                  "Somewhere else (SPECIFY)?" = "Somewhere else")

#Created new variable `race` to combine variables race01:race15 into one column. 
race_data_2 <- race_data %>%
  mutate(race = case_when(race01 == "Yes" ~ "White",
        race02 == "Yes" ~ "Black",
        race03 == "Yes" ~ "Japanese",
        race04 == "Yes" ~ "Chinese",
        race05 == "Yes" ~ "Filipino",
        race06 == "Yes" ~ "Korean",
        race07 == "Yes" ~ "Other Asian or Pacific Islander",
        race08 == "Yes" ~ "American Indian or Alaskan Native",
        race09 == "Yes" ~ "Mexican",
        race10 == "Yes" ~ "Hispanic/Latino",
        race11 == "Yes" ~ "Other",
        race12 == "Yes" ~ "Vietnamese",
        race13 == "Yes" ~ "Asian Indian",
        race14 == "Yes" ~ "Refused",
        race15 == "Yes" ~ "Don't know")) %>%
  select(-(race01:race15))

#Created new variable "packs_per_day" for future calculations for pack-years
smoker_data_2 <- smoker_data %>% mutate(packs_per_day = howmany/20)

```
  
\newpage
  
## Calculating Pack-years
```{r pack-years calculation}
#We calculated pack-years, which is given by the formula of
#pack-years = # of packs of cigarettes smoked per day * years a person has smoked.
#This calculation led to the creation of a new variable called `pack_years`.
#`pack_years` was created conditionally based on the three different time units
#as determined by the existing variable `smok6uni`, which are: "Days", "Months",
#and "Years". To change the unit of "Days" to years, we set up a conditional
#statement in the code to divide `smok6num` by 365 before multiplying the result
#by `packs_per_day` to get pack-years. Similarly, to change the unit of "Months"
#to years, we set up a conditional statement in the code to divide `smok6num`
#by 12 before multiplying the result by `packs_per_day` to get pack-years. For
#`smok6uni` observations that have the value "Years", we just multiplied
#`smok6num` by `packs_per_day` to get pack-years directly. We assigned
#this overall change in the data frame smoker_data_2 to a new data frame
#called smoker_data_3, which includes the new variable `pack_years`.

smoker_data_3 <- smoker_data_2 %>%
  mutate(pack_years =
           case_when(smok6uni == "Days" ~ packs_per_day*(smok6num/365),
                     smok6uni == "Months" ~ packs_per_day*(smok6num/12),
                     smok6uni == "Years" ~ packs_per_day*(smok6num)))

#We then rounded `pack_years` to the nearest whole number for all observations.

smoker_data_3$pack_years <- round(smoker_data_3$pack_years, 0)

```

\newpage

## Joining Cleaned Data Sets Together
```{r joining race_data_2 and smoker_data_3 together}

#In order to join our two cleaned data sets together, we first had to remove the
#strings of 'DIS' and 'STAT' from the `id` column of race_data_2 by using gsub().
#We overwrote these changes in the race_data_2 data frame and viewed these new
#changes to make sure the `id` variable only contains numbers and no
#characters.

race_data_2$id <- gsub('[DISSTAT]', '', race_data_2$id)

#Next, looking at the smoker_data_3 data frame, we see that the `psraid`
#variable contains each study participant's unique ID number, but the variable
#is a numeric data type. On the other hand, `id` from the race_data_2
#data frame is a character data type. We needed to convert `psraid` then from
#character to numeric data type because 1) `psraid` is an identifier rather than
#a numeric value to mathematically manipulate even if it does contain numbers
#and 2) in order to perform a join, the two variables must be the same data
#type.

smoker_data_3$psraid <- as.character(smoker_data_3$psraid)

#Afterward, we performed an inner join between race_data_2 and
#smoker_data_3 by each study participant's unique ID number, which is
#represented by `id` in race_data_2 and `psraid` in smoker_data_3. We
#chose to do an inner join because we wanted to select participants that
#exist in each of our two data sets for our final data frame.
#We assigned this join to a new data frame called joined_smoking_df.

joined_smoking_df <- inner_join(x = race_data_2, y = smoker_data_3,
                                by=c("id" = "psraid"))

```

\newpage

## Visualizations

_**Table: Average Pack-years by Disease Status**_
```{r table of cigarette pack-years by disease}

#Table for avg pack-years per disease for smokers who have a disease
t_avg_pack_years_disease <- joined_smoking_df %>%
  mutate(disease = case_when(asthma == "Yes" ~ "Asthma",
                            heartdis == "Yes" ~ "Heart Disease",
                            diabetes == "Yes" ~ "Diabetes",
                            othmenill == "Yes" ~ "Mental Illness")) %>%
  select(disease, pack_years) %>%
  filter(!is.na(pack_years), !is.na(disease)) %>%
  group_by(disease) %>%
  summarize(avg_pack_years = round(sum(pack_years)/n(), 0))


#Kable table for avg pack-years per disease for smokers who have a disease
#(produced below)
kable(t_avg_pack_years_disease, 
booktabs=T, 
col.names=c("Disease", "Average Pack-years"),  
align='lcccc', 
caption= 'Average Pack-years for Smokers Who Have a Disease') %>%
kable_styling(full_width = T) %>%
kable_styling(latex_options = "hold_position") %>%
footnote(general = "Data Source: 2011 California Smokers Cohort, CA Dept. of Health")

```
_**Interpretation of Average Pack-years by Disease Status Table:**_
\newline
This table demonstrates the average number of pack-years (i.e., average number of cigarette packs smoked per year) per disease type for smokers who reported having asthma, diabetes, heart disease, and/or mental illness in the 2011 California Smokers Cohort study.

Among smokers who have reported having asthma, heart disease, diabetes, and/or mental illness, those with heart disease have the highest number of average pack-years (28), while those with mental illness have the lowest number of average pack-years (17).

\newpage

_**Bar Chart: Average Pack-years by Race and Mental Illness**_
```{r bar graph of cigarette pack-years by race and mental illness, fig.width=8, fig.height=5}

#We first created a subset of the data frame joined_smoking_df for our disease of
#interest, mental illness, called avg_pack_years_race_othmenill. This subset
#includes only the variable of `race` and average values of the variable
#`pack_years` pertaining to mental illness status. The purpose of creating this
#subset is to simplify the process of creating a graph in the next step by
#showing only the relevant information we need.

avg_pack_years_race_othmenill <- joined_smoking_df %>%
  filter(!is.na(pack_years)) %>%
  group_by(race, othmenill) %>%
  summarize(avg_pack_years = sum(pack_years)/n())


#We then created a bar graph representing avg_pack_years_race_othmenill excluding
#NA values in the variables `othmenill` and `avg_pack_years` since we have
#determined that the NA values do not present valuable information for our
#analyses.

avg_pack_years_race_othmenill %>%
  drop_na(c(othmenill, avg_pack_years)) %>%
  ggplot(aes(x = race, y = avg_pack_years)) +
  geom_bar(aes(fill=race), stat="identity", position = "dodge") +
  coord_flip() +
  guides(fill = "none") +
  labs(x = "Race",
       y = "Average Pack-years (Avg. # of Cigarette Packs Smoked per Year)",
  title = "Average Pack-years by Race & Mental Illness Status",
  caption = "Data Source: 2011 California Smokers Cohort, CA Dept. of Health") +
  scale_y_continuous(labels = function(x) format(x,big.mark=",",
                                                     scientific=FALSE)) +
  facet_wrap(~ othmenill, labeller = labeller(othmenill =
                                      c("No" = "Does Not Have Mental Illness",
                                        "Yes" = "Has Mental Illness"))) +
  theme(plot.title = element_text(hjust = 0.5),
        plot.caption = element_text(hjust = 0.5))
       
```
\newline 
\newline 
\newline _**Interpretation of Average Pack-years by Race and Disease Bar Graph:**_
\newline
\newline
This graph exhibits the number of average pack-years (i.e., average number of cigarette packs smoked per year) according to each race category and mental illness status of smokers in the 2011 California Smokers Cohort study.

Among smokers who have reported having no mental illness, those who identified as "Hispanic/Latino" by race appear to have the greatest number of average pack-years, followed closely by "White" and "American Indian or Alaskan Native", compared to other races in the 2011 California Smokers Cohort.

Among smokers who have reported having mental illness, those who identified as "Other" by race appear to have the greatest number of average pack-years, followed closely by "White" and "American Indian or Alaskan Native", compared to other races in the 2011 California Smokers Cohort.

\newpage

## Bar Graph: Cigarette Purchase Location and Mental Illness ##

```{r}
#As similarly performed as the graph above, 
#we also excluded NA values for wherebuy and othmenill variables prior to creating this graph.
#We chose to create a stack bar graph in order to more easily compare the mental illness status 
#reported by smokers for each cigarette purchase location as well as showcase
#the total number of times (frequency) cigarettes were purchased at each location by smokers.

location_othemenill_graph <- joined_smoking_df %>%
  select(wherebuy,othmenill,pack_years) %>%
  filter(!is.na(wherebuy), !is.na(othmenill)) %>%
  ggplot(aes(x=wherebuy)) + 
  geom_bar(aes(fill=othmenill), position="stack") +
  theme_minimal(base_size=6) +
  scale_fill_discrete(name = "Has Mental Illness",) + 
  labs(x="Cigarette Purchase Location",
       y="Frequency of Cigarette Purchase",
       title="Cigarette Purchase Location and Mental Illness Status 
       Among Smokers in 2011 Califoria Smokers Cohort Study")
location_othemenill_graph

```
\newpage

_**Interpretation for Mental Illness Status by Cigarette Purchase Location Bar Graph:**_
\newline
\newline
This bar graph explores the relationship between cigarette purchase location and mental illness status among smokers
in the 2011 California Smokers Cohort study.

Mental illness was not reported by majority of the smokers from each cigarette purchase location. However, mental illness was reported in the greatest number by those who purchased cigarettes at convenience stores or gas stations followed by those who purchased cigarettes at tobacco discount stores, two locations that also have the highest number of cigarette purchases among smokers.
